Detecting diversifying selection for a trait from within and between-species genotypes and phenotypes

### 2024-03-13 manuscript/main-revision.tex
Revision round 1

%! BibTeX Compiler = biber
%TC:ignore
\documentclass{article}
\usepackage{caption}
\usepackage{censor}
\usepackage{xcolor, colortbl}
\definecolor{BLUELINK}{HTML}{0645AD}
\definecolor{DARKBLUELINK}{HTML}{0B0080}
\definecolor{LIGHTGREY}{gray}{0.9}
\PassOptionsToPackage{hyphens}{url}
\usepackage[colorlinks=false]{hyperref}
% for linking between references, figures, TOC, etc in the pdf document
\hypersetup{colorlinks,
    linkcolor=DARKBLUELINK,
    anchorcolor=DARKBLUELINK,
    citecolor=DARKBLUELINK,
    filecolor=DARKBLUELINK,
    menucolor=DARKBLUELINK,
    urlcolor=BLUELINK
} % Color citation links in purple
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{naturalnames}{hyperref}

\usepackage{biorxiv}
\usepackage[backend=biber,eprint=false,isbn=false,url=false,giveninits=true, style=apa,date=year]{biblatex}
\DeclareDelimFormat{nameyeardelim}{\addcomma\space}
\addbibresource{references.bib}

@@

\newcommand{\defEqual}{\stackrel{\text{def}}{=}}
\newcommand{\Multiply}{\cdot}
\newcommand{\MultiplyMatrix}{\times}
\newcommand{\UniDimArray}[1]{\bm{#1}}
\newcommand{\BiDimArray}[1]{\bm{#1}}
\newcommand{\tr}{^{\intercal}}
\newcommand{\inv}{^{-1}}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\Var}{\text{var}}
\DeclareMathOperator{\Cov}{\text{cov}}
\newcommand{\Qst}{Q$_\text{ST}$}
\newcommand{\Fst}{F$_\text{ST}$}
\newcommand{\QstFst}{\Qst--\Fst}
\newcommand{\der}{\mathrm{d}}
\newcommand{\e}{\text{e}}
\newcommand{\Ne}{N_{\text{e}}}
\newcommand{\dnds}{\omega}
\newcommand{\pn}{\pi_N}
\newcommand{\ps}{\pi_S}
\newcommand{\pnps}{\pn / \ps}
\newcommand{\proba}{\mathbb{P}}
\newcommand{\pfix}{\proba_{\text{fix}}}
\newcommand{\Spi}{i}
\newcommand{\Spj}{j}
\newcommand{\NbrTaxa}{n}
\newcommand{\Time}{T}
\newcommand{\GenTime}{\tau}
\newcommand{\NbrGen}{t_{\Spi, \Spj}}
\newcommand{\NucDiv}{d_{\Spi, \Spj}}
\newcommand{\Trait}{P}
\newcommand{\Heritability}{h^2}
\newcommand{\HeritabilitySpi}{\Heritability_{\Spi}}
\newcommand{\MeanTrait}{\bar{\Trait}}
\newcommand{\VecTrait}{\UniDimArray{\bar{\Trait}}}
\newcommand{\RootTrait}{\theta}
\newcommand{\VarPhy}{\Cov \left( \MeanTrait_{\Spi}, \MeanTrait_{\Spj}\right)}
\newcommand{\VecZero}{\UniDimArray{0}}
\newcommand{\VecOne}{\UniDimArray{1}}
\newcommand{\Distance}{\BiDimArray{D}}
\newcommand{\DistanceMatrix}{\BiDimArray{\Distance}}
\newcommand{\MutationRatePheno}{\mu}
\newcommand{\MutationRateNuc}{u}
\newcommand{\SubRate}{q}
\newcommand{\NbrLoci}{L}
\newcommand{\VarPhenotype}{V_{\Trait}}
\newcommand{\VarPhenotypeSpi}{V_{\Trait, \Spi}}
\newcommand{\VarGenetic}{V_{\mathrm{A}}}
\newcommand{\VarGeneticSpi}{V_{\mathrm{A}, \Spi}}
\newcommand{\VarEnv}{V_{\mathrm{E}}}
\newcommand{\VarMutation}{V_{\mathrm{M}}}
\newcommand{\GenArchi}{\NbrLoci \Multiply \E \left[ a^2 \right]}
\newcommand{\RateBetween}{\sigma^2_{\mathrm{B}}}
\newcommand{\RateWhithin}{\sigma^2_{\mathrm{W}}}
\newcommand{\RateWhithinSpi}{\sigma^2_{\mathrm{W}, \Spi}}
\newcommand{\VecRateWhithin}{\UniDimArray{\RateWhithin}}
\newcommand{\EstRateBetween}{\widehat{\RateBetween}}
\newcommand{\EstRateWhithin}{\widehat{\RateWhithin}}
\newcommand{\NI}{\rho}
\newcommand{\EstNI}{\widehat{\rho}}
\newcommand{\StdSelection}{\sigma}
\newcommand{\VarSelection}{\StdSelection^2}

@@

% Research Article (7500 words)
\section{Introduction}\label{sec:introduction}

Determining whether a trait is under a particular regime of selection has been a long-standing goal in evolutionary biology.
Fundamentally, distinguishing neutral evolution from selection requires determining which selective regime is supported by the observed variation of traits or sequences.
The variation of phenotypes (traits) and genotypes (sequences) can be observed at different scales, across different development stages at the individual level, across different individuals and populations at the species level, and finally across different species at the phylogenetic level.
All these systems require different assumptions and methodologies, and the endeavor to determine the selective regime for a given trait has thus incorporated theories, methods, and developments across various fields of evolutionary biology such as quantitative genetics, population genetics, phylogenetics and comparative genomics~\parencite{lynch_genetics_1998, walsh_evolution_2018}.

Leveraging individual variations within the same species, Genome-Wide Association Studies (GWAS) in humans have shown that traits are mostly polygenic (many loci associated with a given trait) and under stabilizing selection, while the loci affecting those traits are mostly pleiotropic (many traits associated with a given locus) with additive effects~\parencite{simons_population_2018, sella_thinking_2019}.
Across several populations, by contrasting both trait and genetic differentiation, \QstFst\ methods have been used to determine the selective regime and to quantify the strength of selection acting on a trait~\parencite{merila_comparison_2001, leinonen_comparative_2008}.
A trait differentiation (\Qst) higher than genetic differentiation (\Fst) is interpreted as a signature of diversifying selection due to adaptation to different optimum trait values in the different populations.
Contrarily, \Qst\ lower than \Fst\ is interpreted as a signature of stabilizing selection~\parencite{lamy_qst_2012}.
However, it has been found that \QstFst\ methods require many populations~\parencite{ohara_bias_2005}, and that various factors can generate a spurious signal of selection~\parencite{pujol_are_2008, edelaar_comparisons_2011}.
Alternatively, methods derived the statistical distribution of population means generated by random genetic drift and used the probability density of this distribution to test whether the observed pattern could be generated by drift alone~\parencite{ovaskainen_new_2011}.
Moreover, the test for diversifying selection is limited to recent local adaptation since the test is based on the variation observed within a single species.

% Pair of species
To disentangle selection from neutral evolution, trait variation can also be observed at a larger time scale.
For example, starting from the same ancestral population, divergent lineages accumulate phenotypic changes reaching fixation in the population.
These changes ultimately result in different mean trait values across lineages.
Theoretically, the variance in mean trait value (between lineages) does increase linearly with time of divergence, and also proportionally to the trait variance at the population scale~\parencite{lande_genetic_1980, turelli_heritable_1984}.
Empirically, this effect can be observed for genes with larger within-species variation in gene expression level, which exhibits a faster accumulation of divergence in mean expression level~\parencite{khaitovich_neutral_2004}.
Analogously, in the context of protein-coding DNA sequences, leveraging within species variation and divergence to a sister species is the crux of the \textcite{mcdonald_adaptative_1991} test.
In such a test, inflation of divergence to the sister species is compared to polymorphism within species, while neutral makers (usually synonymous sites) are used to determine the neutral expectation and thus are used for normalization.
Altogether, both the trait variance and the evolution in mean value can be used to test for trait selection in a pair of species~\parencite{walsh_evolution_2018}.

% Phylogenetics (mean trait evolution)
Alternatively, by accounting for the underlying relationships between several species, the selective regime for a quantitative trait can also be tested at the phylogenetic scale~\parencite{felsenstein_phylogenies_1985}.
Under neutral evolution, the change in mean trait value along a given branch of the tree is normally distributed, with a variance proportional to divergence time~\parencite{hansen_translating_1996}.
As a result, the mean trait value can be modeled as a Brownian process branching at every node of the tree~\parencite{hansen_translating_1996, harmon_phylogenetic_2018}.
Reconstructing the trait variation along the whole phylogeny as a Brownian process can thus constitute a null model of neutral trait evolution.
Deviations from the assumptions of the Brownian process are however well known.
When trait variation is constrained because of optimum mean trait values across or between species, the pattern of evolution can be modeled by the Ornstein-Uhlenbeck processes, which is often interpreted as a signature of stabilizing selection~\parencite{catalan_drift_2019}.
Alternatively, a trend in the Brownian process (the tendency of a trait to evolve in a certain direction without fixed optimum) is interpreted as a signature of directional selection at the phylogenetic scale~\parencite{silvestro_early_2019}.
However, studies have shown that such comparative approaches are subject to different biases~\parencite{harmon_phylogenetic_2018}.
First, a trait under stabilizing selection for which the optimal trait value is also changing as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral~\parencite{hansen_translating_1996}.
In other words, the better fit of a Brownian process does not necessarily constitute proof of the neutral model.
Also, a better fit of a Brownian could be due to a trait evolving with a rate two low compared to the timespan on which it is measured~\parencite{grabowski_cautionary_2023}.
Second, even for a trait evolving under a neutral regime, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts~\parencite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
Those limitations, altogether with the use of mean trait estimates leaving out the variance in traits between individuals, easily generate misclassification of selection from methods at the phylogenetic scale.

% At the frontier Phylogenetics/Population-genetics
At the frontier between micro and macro-evolution, comparative methods at the phylogenetic scale have acknowledged the importance of modeling within-species variation together with changes in mean trait value to either describe measurement errors~\parencite{lynch_methods_1991, hansen_interpreting_2012}, incorporate values for individuals~\parencite{felsenstein_comparative_2008} or to scale the rate of change in mean trait value~\parencite{kostikova_bridging_2016, gaboriau_multiplatform_2020, gaboriau_exploring_2023}.
Between to within-species variation has also been used to infer diversifying selection by estimating the ratio of between to within variation of many traits and test for deviation from the average ratio across traits~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
Here, our goal was again to use both variances between and within species to determine the selective regime of a quantitative trait.
We build a novel framework that integrates trait variation at the phylogenetic and population scales together with estimates of nucleotide sequence variations at both scales.
It allowed us to define an expected ratio of normalized variance between and within species while setting the threshold of this ratio for neutral, stabilizing, and diversifying selection.
The ratio that we propose can be considered as a neutrality index for a any quantitative trait articulating trait and nucleotide variation within and between species.
Importantly, our neutrality index also leverages nucleotide divergence and polymorphism to normalize trait variation at both scales, such that it does not require estimating population size (within-species) or speciation time (between species).
From the field of population genetics, our study can be seen as the macro-evolutionary generalization of \QstFst\ methods to account for phylogenetic relationships between species~\parencite{ovaskainen_new_2011}.
From the field of phylogenetics, our study can be seen as an alternative to the EVE model~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015} for a single trait, where we set a threshold for neutral evolution by leveraging species nucleotide polymorphism and divergence.
% Basically, we are testing whether the variance of means is greater than the mean of variances

\section{Materials and Methods}\label{sec:materials-and-methods}
\subsection{Neutrality index for a quantitative trait}\label{subsec:neutrality-index-for-a-quantitative-trait}

While observing trait variations across individuals of several species, we ask if the variation within species compared to variation between species is compatible with neutral evolution or not.
In statistical terms, this can also be framed as: Is the variance of means equal to the mean of variances?
The difficulty in such a study is that individuals are not independent samples, but are from species that diverged at different times, and each species has a putatively different population size.
By reviewing theoretical expectations and leveraging nucleotide sequence variations, the goal of this section is thus to obtain normalized trait variation between and within species that are equal if the trait is neutral.
Here we denote these normalized trait variations as respectively $\RateWhithin$ for within species and as $\RateBetween$ for between species.

\subsubsection{Within-species trait variations}

For a given trait, the genetic architecture is mainly defined by the number of loci encoding the trait ($\NbrLoci$) and the random additive effect of a mutation on the trait ($a$).
For a diploid individual, the mutational variance ($\VarMutation$) is the rate at which new mutations contribute to the trait variance per generation.
As shown in \textcite{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ is a function of the mutation rate per loci per generation ($\MutationRatePheno$) and the genetic architecture of the trait as:
\begin{gather}
    \VarMutation = 2 \MutationRatePheno \Multiply \GenArchi \label{eq:var-mutation}.
\end{gather}

While in an infinitesimal model mutations supply new genetic variants, random genetic drift depletes standing variation~\parencite{turelli_commentary_2017, barton_infinitesimal_2017, sella_thinking_2019}.
For a neutral trait at equilibrium between mutation and drift~\parencite{lynch_mutation_1998}, the additive genetic variance in a species ($\VarGenetic$) is a function of the mutational variance ($\VarMutation$) and the effective number of individuals in the population ($\Ne$):
\begin{align}
    \VarGenetic & =  2 \Ne \Multiply \VarMutation, \\
    & = 4 \Ne \Multiply \MutationRatePheno \Multiply \GenArchi \text{ from eq.~\ref{eq:var-mutation}}\label{eq:var-genetic}.
\end{align}

For any neutral genomic region of interest, the nucleotide diversity, $\pi$, is the average number of differences between pairs of sequences drawn at random, which is also equal to the sum of expected heterozygosities over all loci~\parencite{tajima_statistical_1989}.
Any segregating mutations will eventually reach fixation or extinction due to random genetic drift and $\pi$ is also at a balance between mutations and drift.
As shown in \textcite{tajima_statistical_1989}, $\pi$ is a function of the mutation rate per loci per generation ($\MutationRateNuc$) and the effective population size ($\Ne$):
\begin{gather}
    \pi = 4 \Ne \Multiply \MutationRateNuc \label{eq:genetic-diversity}.
\end{gather}

We define $\RateWhithin$ as the ratio of additive genetic variance of the trait ($\VarGenetic$) over $\pi$ of any neutral genomic region of interest.
This ratio allows removing the effect of $\Ne$, giving $\RateWhithin$ as a proxy of the underlying genetic architecture:
\begin{align}
    \RateWhithin & \defEqual \frac{\VarGenetic }{\pi} \label{eq:def-rate-pop}, \\
    & = \frac{4 \Ne \Multiply \MutationRatePheno \Multiply \GenArchi}{4 \Ne \Multiply \MutationRateNuc} \text{ from eq.~\ref{eq:var-mutation} and~\ref{eq:genetic-diversity}}, \\
    & = \frac{\MutationRatePheno \Multiply \GenArchi}{\MutationRateNuc}. \label{eq:rate-pop}
\end{align}

The additive genetic variance is also equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by narrow-sense heritability ($\Heritability$; \parencite{hill_data_2008}), as:
\begin{align}
    \VarGenetic & = \Heritability \Multiply \VarPhenotype.\label{eq:var-gen-pheno}
\end{align}
Which leads to $\RateWhithin$ being a function of $\VarPhenotype$ and $\Heritability$ as:
\begin{align}
    \RateWhithin & = \frac{\Heritability \Multiply \VarPhenotype }{\pi } \text{ from definition eq.~\ref{eq:def-rate-pop} and~\ref{eq:var-gen-pheno}}, \label{eq:def-rate-pheno-pop} \\
    & = \frac{\MutationRatePheno \Multiply \GenArchi}{\MutationRateNuc} \text{ from eq.~\ref{eq:rate-pop}} \label{eq:rate-pheno-pop}
\end{align}

>>>>

%! BibTeX Compiler = bibtex
%TC:ignore
\documentclass{article}
\usepackage{caption}
\usepackage{censor}
\usepackage{xcolor, colortbl}
\definecolor{BLUELINK}{HTML}{0645AD}
\definecolor{DARKBLUELINK}{HTML}{0B0080}
\definecolor{LIGHTGREY}{gray}{0.9}
\PassOptionsToPackage{hyphens}{url}
\usepackage[colorlinks=false]{hyperref}
% for linking between references, figures, TOC, etc in the pdf document
\hypersetup{colorlinks,
    linkcolor=DARKBLUELINK,
    anchorcolor=DARKBLUELINK,
    citecolor=DARKBLUELINK,
    filecolor=DARKBLUELINK,
    menucolor=DARKBLUELINK,
    urlcolor=BLUELINK
} % Color citation links in purple
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{naturalnames}{hyperref}

\usepackage{biorxiv}

@@

\newcommand{\defEqual}{\stackrel{\text{def}}{=}}
\newcommand{\Multiply}{\cdot}
\newcommand{\MultiplyMatrix}{\times}
\newcommand{\UniDimArray}[1]{\bm{#1}}
\newcommand{\BiDimArray}[1]{\bm{#1}}
\newcommand{\tr}{^{\intercal}}
\newcommand{\inv}{^{-1}}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\Var}{\text{var}}
\DeclareMathOperator{\Cov}{\text{cov}}
\newcommand{\Qst}{Q$_\text{ST}$}
\newcommand{\Fst}{F$_\text{ST}$}
\newcommand{\QstFst}{\Qst--\Fst}
\newcommand{\der}{\mathrm{d}}
\newcommand{\e}{\text{e}}
\newcommand{\Ne}{N_{\text{e}}}
\newcommand{\dnds}{\omega}
\newcommand{\pn}{\pi_N}
\newcommand{\ps}{\pi_S}
\newcommand{\pnps}{\pn / \ps}
\newcommand{\proba}{\mathbb{P}}
\newcommand{\pfix}{\proba_{\text{fix}}}
\newcommand{\Indiv}{k}
\newcommand{\Branch}{b}
\newcommand{\WishartIDD}{m}
\newcommand{\Spi}{i}
\newcommand{\Spj}{j}
\newcommand{\NbrTaxa}{n}
\newcommand{\Time}{t}
\newcommand{\NbrGen}{t_{\Spi, \Spj}}
\newcommand{\NucDiv}{d_{\Spi, \Spj}}
\newcommand{\Trait}{P}
\newcommand{\Heritability}{h^2}
\newcommand{\HeritabilitySpi}{\Heritability_{\Spi}}
\newcommand{\MeanTrait}{\bar{\Trait}}
\newcommand{\VecTrait}{\UniDimArray{\bar{\Trait}}}
\newcommand{\RootTrait}{\phi}
\newcommand{\VarPhy}{\Cov \left( \MeanTrait_{\Spi}, \MeanTrait_{\Spj}\right)}
\newcommand{\VecZero}{\UniDimArray{0}}
\newcommand{\VecOne}{\UniDimArray{1}}
\newcommand{\Distance}{\BiDimArray{D}}
\newcommand{\DistanceMatrix}{\BiDimArray{\Distance}}
\newcommand{\MutationRatePheno}{\mu}
\newcommand{\MutationRateNuc}{u}
\newcommand{\SubRate}{q}
\newcommand{\NbrLoci}{L}
\newcommand{\VarPhenotype}{V_{\Trait}}
\newcommand{\VarPhenotypeSpi}{V_{\Trait, \Spi}}
\newcommand{\VarGenetic}{V_{\mathrm{A}}}
\newcommand{\VarGeneticSpi}{V_{\mathrm{A}, \Spi}}
\newcommand{\VarEnv}{V_{\mathrm{E}}}
\newcommand{\VarMutation}{V_{\mathrm{M}}}
\newcommand{\GenArchi}{\NbrLoci \Multiply \E \left[ a^2 \right]}
\newcommand{\RateMut}{\sigma^2_{\mathrm{M}}}
\newcommand{\RateBetween}{\sigma^2_{\mathrm{B}}}
\newcommand{\RateWhithin}{\sigma^2_{\mathrm{W}}}
\newcommand{\RateWhithinSpi}{\sigma^2_{\mathrm{W}, \Spi}}
\newcommand{\VecRateWhithin}{\UniDimArray{\RateWhithin}}
\newcommand{\EstRateBetween}{\widehat{\sigma}^2_{\mathrm{B}}}
\newcommand{\EstRateWhithin}{\widehat{\sigma}^2_{\mathrm{W}}}
\newcommand{\NI}{\rho}
\newcommand{\EstNI}{\widehat{\rho}}
\newcommand{\StdSelection}{\sigma}
\newcommand{\VarSelection}{\StdSelection^2}

@@

% Research Article (7500 words)
\section*{Introduction}\label{sec:introduction}

Determining whether a trait is under a particular regime of selection has been a long-standing goal in evolutionary biology.
Fundamentally, distinguishing neutral evolution from selection requires determining which selective regime is supported by the observed variation of traits or sequences.
The variation of phenotypes (traits) and genotypes (sequences) can be observed at different scales, across different development stages at the individual level, across different individuals and populations at the species level, and finally across different species at the phylogenetic level.
All these systems require different assumptions and methodologies, and the endeavor to determine the selective regime for a given trait has thus incorporated theories, methods, and developments across various fields of evolutionary biology such as quantitative genetics, population genetics, phylogenetics and comparative genomics~\citep{lynch_genetics_1998, walsh_evolution_2018}.

Leveraging individual variations within the same species, Genome-Wide Association Studies (GWAS) in humans have shown that traits are mostly polygenic (many loci associated with a given trait) and under stabilizing selection, while the loci affecting those traits are mostly pleiotropic (many traits associated with a given locus) with additive effects~\citep{simons_population_2018, sella_thinking_2019}.
Given this genetic architecture of traits, from two diverging populations, it is possible to distinguish which traits have evolved under natural selection in controlled experimental settings, by performing genetic cross between individuals~\citep{fraser_detecting_2020}.
Across several populations, by contrasting both trait differentiation (\Qst) and genetic differentiation (\Fst), so-called \QstFst\ methods have been used to determine the selective regime and to quantify the strength of selection acting on a trait~\citep{merila_comparison_2001, leinonen_comparative_2008}.
\Qst\ higher than \Fst\ is interpreted as a signature of diversifying selection due to adaptation to different optimum trait values in the different populations.
Contrarily, \Qst\ lower than \Fst\ is interpreted as a signature of stabilizing selection~\citep{lamy_qst_2012}.
Other frameworks explicitly model genetic drift as a random process generating both trait and genetic differences between individuals and populations.
This integrated framework can discriminate between selection and genetic drift as a cause of trait differentiation between populations of the same species~\citep{ovaskainen_new_2011}.
However, regardless of the strengths and weaknesses of each method\citep{pujol_are_2008, edelaar_comparisons_2011, ovaskainen_new_2011}, tests of trait differentiation between populations are ultimately limited to recent local adaptation since they are based on the variation observed within a single species.

% Pair of species
To disentangle selection from neutral evolution, trait variation can also be observed at a larger time scale.
For example, starting from the same ancestral population, divergent lineages accumulate phenotypic changes that will reach fixation in the population.
These changes ultimately result in different mean trait values across lineages.
Theoretically, the variance in mean trait value (between lineages) does increase linearly with time of divergence, and also proportionally to the trait variance at the population scale~\citep{lande_genetic_1980, turelli_heritable_1984}.
Empirically, this effect can be observed for genes with larger within-species variation in gene expression level, which exhibits a faster accumulation of divergence in mean expression level~\citep{khaitovich_neutral_2004}.
As an analogy, in the context of protein-coding DNA sequences, leveraging within species variation and divergence to a sister species is the crux of the \citet{mcdonald_adaptative_1991} test.
In such a test, inflation of divergence to the sister species is compared to polymorphism within species, while neutral makers (usually synonymous sites) are used to determine the neutral expectation and thus are used for normalization.
Altogether, both the trait variance and the evolution in mean value can be used to test for trait selection in a pair of species~\citep{walsh_evolution_2018}.

% Phylogenetics (mean trait evolution)
Alternatively, by accounting for the underlying relationships between several species, the selective regime for a quantitative trait can also be tested at the phylogenetic scale~\citep{felsenstein_phylogenies_1985}.
Under neutral evolution, the change in mean trait value along a given branch of the tree is normally distributed, with a variance proportional to divergence time~\citep{hansen_translating_1996}.
As a result, the mean trait value can be modeled as a Brownian process branching at every node of the tree~\citep{hansen_translating_1996, harmon_phylogenetic_2018}.
Reconstructing the trait variation along the whole phylogeny as a Brownian process can thus constitute a null model of neutral trait evolution.
Deviations from the assumptions of the Brownian process are however well known.
When trait variation is constrained because of optimum mean trait values across or between species, the pattern of evolution can be modeled by the Ornstein-Uhlenbeck processes, which is often interpreted as a signature of stabilizing selection~\citep{catalan_drift_2019}.
Alternatively, a trend in the Brownian process (the tendency of a trait to evolve in a certain direction without fixed optimum) is interpreted as a signature of directional selection at the phylogenetic scale~\citep{silvestro_early_2019}.
However, studies have shown that such comparative approaches are subject to different biases~\citep{harmon_phylogenetic_2018}.
First, a trait under stabilizing selection for which the optimal trait value is also changing as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral~\citep{hansen_translating_1996}.
In other words, the better fit of a Brownian process does not necessarily constitute proof of the neutral model.
Also, a better fit of a Brownian could be due to a trait evolving with a rate two low compared to the timespan on which it is measured~\citep{grabowski_cautionary_2023}.
Second, even for a trait evolving under a neutral regime, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts~\citep{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
Those limitations, altogether with the use of mean trait estimates leaving out the variance in traits between individuals, easily generate misclassification of selection from methods at the phylogenetic scale.

% At the frontier Phylogenetics/Population-genetics
At the frontier between micro and macro-evolution, comparative methods at the phylogenetic scale have acknowledged the importance of modeling within-species variation together with changes in mean trait value to either describe measurement errors~\citep{lynch_methods_1991, hansen_interpreting_2012}, incorporate values for individuals~\citep{felsenstein_comparative_2008} or to scale the rate of change in mean trait value~\citep{kostikova_bridging_2016, gaboriau_multiplatform_2020, gaboriau_exploring_2023}.
Between to within-species variation has also been used to infer diversifying selection by estimating the ratio of between to within variation of many traits and test for deviation from the average ratio across traits~\citep{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
Here, our goal was again to use both variances between and within species to determine the selective regime of a quantitative trait.
We build a novel framework that integrates trait variation at the phylogenetic and population scales together with estimates of nucleotide sequence variations at both scales.
It allowed us to define an expected ratio of normalized variance between and within species while setting the threshold of this ratio for neutral, stabilizing, and diversifying selection.
The ratio that we propose can be considered as a neutrality index for a any quantitative trait articulating trait and nucleotide variation within and between species.
Importantly, our neutrality index also leverages nucleotide divergence and polymorphism to normalize trait variation at both scales, such that it does not require estimating population size (within-species) or speciation time (between species).
From the field of population genetics, while \QstFst\ methods and their derivatives ultimately seek trait differentiation among different populations from the same species~\citep{pujol_are_2008, ovaskainen_new_2011}, our study can be seen as their macro-evolutionary generalization to account for phylogenetic relationships between species.
From the field of phylogenetics, our study can be seen as an alternative to the EVE model~\citep{rohlfs_modeling_2014, rohlfs_phylogenetic_2015} for a single trait, where we set a threshold for neutral evolution by leveraging species nucleotide polymorphism and divergence.

\section*{Materials and Methods}\label{sec:materials-and-methods}
\subsection*{Neutrality index for a quantitative trait}\label{subsec:neutrality-index-for-a-quantitative-trait}

While observing trait variations across individuals of several species, we ask if the variation within species compared to variation between species is compatible with neutral evolution or not.
In statistical terms, this can also be framed as: Is the variance of means equal to the mean of variances?
The difficulty in such a study is that individuals are not independent samples, but are from species that diverged at different times.
By reviewing theoretical expectations and leveraging nucleotide sequence variations, the goal of this section is thus to obtain normalized trait variation between and within species that are equal if the trait is neutral.
Here we denote these normalized trait variations as respectively $\RateWhithin$ for within species and as $\RateBetween$ for between species.

\subsubsection*{Within-species trait variations}

For a given trait, the genetic architecture is mainly defined by the number of loci encoding the trait ($\NbrLoci$) and the random additive effect of a mutation on the trait ($a$).
For a diploid individual, the mutational variance ($\VarMutation$) is the rate at which new mutations contribute to the trait variance per generation.
As shown in \citet{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ is a function of the mutation rate per locus per generation ($\MutationRatePheno$) and the genetic architecture of the trait as:
\begin{gather}
    \VarMutation = 2 \MutationRatePheno \Multiply \GenArchi \label{eq:var-mutation}.
\end{gather}

While in an infinitesimal model mutations supply new genetic variants, random genetic drift depletes standing variation~\citep{turelli_commentary_2017, barton_infinitesimal_2017, sella_thinking_2019}.
For a neutral trait at equilibrium between mutation and drift~\citep{lynch_mutation_1998}, the additive genetic variance in a species ($\VarGenetic$) is a function of the mutational variance ($\VarMutation$) and the effective number of individuals in the population ($\Ne$):
\begin{align}
    \VarGenetic & =  2 \Ne \Multiply \VarMutation, \\
    & = 4 \Ne \Multiply \MutationRatePheno \Multiply \GenArchi \text{ from eq.~\ref{eq:var-mutation}}\label{eq:var-genetic}.
\end{align}

For any neutral genomic region of interest, the nucleotide diversity, $\pi$, is the average number of differences between pairs of sequences drawn at random, which is also equal to the sum of expected heterozygosities over all nucleotide sites~\citep{tajima_statistical_1989}.
Any segregating mutations will eventually reach fixation or extinction due to random genetic drift and $\pi$ is also at a balance between mutations and drift.
As shown in \citet{tajima_statistical_1989}, $\pi$ is a function of the mutation rate ($\MutationRateNuc$, per nucleotide site per generation) and the effective population size ($\Ne$):
\begin{gather}
    \pi = 4 \Ne \Multiply \MutationRateNuc \label{eq:genetic-diversity}.
\end{gather}

To remove the effect of $\Ne$, we define $\RateWhithin$ as the ratio of additive genetic variance of the trait ($\VarGenetic$) over $\pi$ of any neutral genomic region of interest.
After simplification, $\RateWhithin$ is then solely a function of the underlying genetic architecture as:
\begin{align}
    \RateWhithin & \defEqual \frac{\VarGenetic }{\pi} \label{eq:def-rate-pop}, \\
    & = \frac{4 \Ne \Multiply \MutationRatePheno \Multiply \GenArchi}{4 \Ne \Multiply \MutationRateNuc} \text{ from eq.~\ref{eq:var-mutation} and~\ref{eq:genetic-diversity}}, \\
    & = \frac{\MutationRatePheno \Multiply \GenArchi}{\MutationRateNuc}. \label{eq:rate-pop}
\end{align}

If $\VarGenetic$ is not empirically accessible, it can be related to the observed phenotypic variance ($\VarPhenotype$), multiplied by narrow-sense heritability of the trait ($\Heritability$), as~\citep{hill_data_2008}:
\begin{align}
    \VarGenetic & = \Heritability \Multiply \VarPhenotype.\label{eq:var-gen-pheno}
\end{align}
Which leads to $\RateWhithin$ being a function of $\VarPhenotype$ and $\Heritability$ instead of $\VarGenetic$ as:
\begin{align}
    \RateWhithin & = \frac{\Heritability \Multiply \VarPhenotype }{\pi } \text{ from definition eq.~\ref{eq:def-rate-pop} and~\ref{eq:var-gen-pheno}}, \label{eq:def-rate-pheno-pop} \\
    & = \frac{\MutationRatePheno \Multiply \GenArchi}{\MutationRateNuc} \text{ from eq.~\ref{eq:rate-pop}} \label{eq:rate-pheno-pop}
\end{align}

\subsubsection*{Between-species trait variations}\label{subsec:between-species-var}
For a given species $\Spi$, we denote by $\MeanTrait_{\Spi}$ the mean value of the trait across the individuals.
If the trait is neutral and encoded by many loci as assumed by the infinitesimal model, $\MeanTrait_{\Spi}$ evolves as a Brownian process~\citep{felsenstein_phylogenies_1985, hansen_translating_1996}.
Given a phylogenetic tree, for a pair of species $\Spi$ and $\Spj$ from this tree, we denote as $\NbrGen$ the number of generations between the root of the tree and the most recent common ancestor of taxa $\Spi$ and $\Spj$.
Then, the covariance between $\MeanTrait_{\Spi}$ and $\MeanTrait_{\Spj}$ depends on $\NbrGen$ as given in \citet{hansen_translating_1996} by:
\begin{align}
    \VarPhy & = \frac{\VarGenetic}{\Ne} \Multiply \NbrGen \\
    & = 4 \NbrGen \Multiply \MutationRatePheno \Multiply \GenArchi \label{eq:covar},\text{ from eq.~\ref{eq:var-genetic}}.
\end{align}
Moreover, for any genomic region under neutral evolution, some mutations will eventually reach fixation due to random genetic drift, resulting in a substitution of a nucleotide at the species level.
The probability of fixation ($\pfix$) of a neutral mutation is $1/2\Ne$~\citep{kimura_probability_1962}.
We can derive the substitution rate ($\SubRate$, per nucleotide site per generation) as the number of newly arisen mutations ($2\Ne \Multiply \MutationRateNuc$) multiplied by the probability of fixation for each newly arisen mutations $\pfix$~\citep{kimura_evolutionary_1968}, giving:
\begin{align}
    \SubRate & = 2 \Ne \Multiply \MutationRateNuc \Multiply \pfix, \\
    & = 2 \Ne  \Multiply \MutationRateNuc  \Multiply \dfrac{1}{2\Ne}, \\
    & = \MutationRateNuc. \label{eq:substitution-rate}
\end{align}
That is, if mutations are neutral, the rate of substitution per generation within a genomic region equals the rate at which new mutations arise per generation for the same genomic region~\citep[review]{mccandlish_modeling_2014}.
